# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
```{r}

setwd('C:\\Users\\Kingsbury\\Documents\\GitHub\\RepData_PeerAssessment1')

# DataLoad
activity <- read.csv('activity.csv',
                                header=TRUE, sep=",", na.strings = "NA")

## Convert the Date column to date format
activity$date <-  as.Date(activity$date,format="%Y-%m-%d")

#show the table of NAs
table(is.na(activity$steps))

```
## What is mean total number of steps taken per day?
```{r}

#calc the total nbr steps taken per day
total_nbr_steps_day <- aggregate(activity$steps, by = list(activity$date), FUN = sum)
#rename columns
colnames(total_nbr_steps_day) <- c("date", "nbr_steps")

#calculate mean and median,  remove NAs
mean_raw <- mean(total_nbr_steps_day$nbr_steps, na.rm = TRUE)
median_raw <- median(total_nbr_steps_day$nbr_steps, na.rm = TRUE)

```

```{r}
# print the values to screen 
mean_raw
median_raw
```

### Histogram of total number of steps taken per day
```{r historgram_tot_nbr_steps_day, fig.width=7, fig.height=6, echo=TRUE}
# draw the histogram
hist(total_nbr_steps_day$nbr_steps[is.na(total_nbr_steps_day$nbr_steps)== FALSE], main="Histogram of tot nbr steps per day", xlab="Nbr steps per day")
rug(total_nbr_steps_day$nbr_steps[is.na(total_nbr_steps_day$nbr_steps)== FALSE], ticksize = 0.02)

```

## What is the average daily activity pattern?
### Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
#removing NAs
clean_activity <-subset(activity, is.na(activity$steps)== FALSE)
# calculating 
clean_activity$avg_by_date <- ave(clean_activity$steps, clean_activity$date)
# draw the time series plot
```

```{r Avg_nbr_steps_per_5m_interval, fig.width=7, fig.height=6, echo=TRUE}
plot(clean_activity$date, clean_activity$avg_by_date,  type="l", ylab="Avg nbr steps per 5m interval", xlab="Date")

```
### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
# calculate average nbr steps per interval
clean_activity$avg_by_interval <- ave(clean_activity$steps, clean_activity$interval)
# plot average nbr steps per interval
#show maximum
max( clean_activity[, "avg_by_interval" ] ) 
# which interval contains the maximum
with(clean_activity,interval[avg_by_interval== max(avg_by_interval)])[1]

```

```{r Average_nbr_of_steps_per_five_min_interval, fig.width=7, fig.height=6, echo=TRUE}
plot(clean_activity$interval, clean_activity$avg_by_interval, ylab="Avg nbr steps", xlab="Interval", main="Average nbr of steps per five min interval")

```

## Imputing missing values

### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
# report NAs
colSums(is.na(activity))

```
### New data set with imputed values
```{r}
#load a time series library zoo
library(zoo)
# convert the data frame into a time series object for usage in zoo
activity_imputed <- read.zoo(activity, index.column = 2, format ="%Y-%m-%d" )

# fill NAs be using median of the five minutes intervals in the zoo ts object
activity_filled <- na.aggregate(activity_imputed, by = 3, FUN = median)

## helper function to convert a time series object to a dataframe
zoo.to.data.frame <- function(x, index.name="date") {
        stopifnot(is.zoo(x))
        xn <- if(is.null(dim(x))) deparse(substitute(x)) else colnames(x)
        setNames(data.frame(index(x), x, row.names=NULL), c(index.name,xn))
}

# convert ts object back to a dataframe
activity_filled_df <- zoo.to.data.frame(activity_filled)
# check NAs
table(is.na(activity_filled_df$steps))

```
### Calculations of median and mean with new NAs filled dataset
```{r}

#calc the total nbr steps taken per day
total_nbr_steps_day_fill <- aggregate(activity_filled_df$steps, by = list(activity_filled_df$date), FUN = sum)
#rename columns
colnames(total_nbr_steps_day_fill) <- c("date", "nbr_steps")

#calculate mean and median,  remove NAs
mean_na_filled <- mean(total_nbr_steps_day_fill$nbr_steps, na.rm = TRUE)
median_na_filled <- median(total_nbr_steps_day_fill$nbr_steps, na.rm = TRUE)

```
### Do these values differ from the estimates from the first part of the assignment? 

```{r}
mean_raw - mean_na_filled
median_raw - median_na_filled

## Yes, there is a difference between mean/median before and after filling of NAs
```

### Histogram of total number of steps taken per day
```{r Histogram_NA_filled, fig.width=7, fig.height=6, echo=TRUE}
# draw histogram
hist(total_nbr_steps_day_fill$nbr_steps[is.na(total_nbr_steps_day_fill$nbr_steps)== FALSE], main="NAs filled: histogram of tot nbr steps per day", xlab="Nbr steps per day")
rug(total_nbr_steps_day_fill$nbr_steps[is.na(total_nbr_steps_day_fill$nbr_steps)== FALSE], ticksize = 0.02)

```

## Are there differences in activity patterns between weekdays and weekends?
### Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day

```{r}
#create a new column with weekdays
activity_filled_df$weekdays <- weekdays(activity_filled_df$date)

#create a new column with two values weekday and weekend
activity_filled_df[((activity_filled_df[,4] == "Saturday")|(activity_filled_df[,4]=="Sunday")),'wd'] <- "weekend"
activity_filled_df[(!((activity_filled_df[,4] == "Saturday")|(activity_filled_df[,4]=="Sunday"))),'wd'] <- "weekday"
#new column with averages per 5 min interval
activity_filled_df$ave <- ave(activity_filled_df$step, activity_filled_df$interval, activity_filled_df$wd)

```
### Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken 
```{r Lattice_timeseries_panel_plot, fig.width=7, fig.height=6, echo=TRUE}

library(lattice)

xyplot(activity_filled_df$ave ~ activity_filled_df$interval | activity_filled_df$wd,
       layout =c(1,2), xlab='inteval', ylab='Number of steps', type ="l")

```

